---
title: "Hands-on Exercise 2"
author: "Nor Hendra"
date: "23 April 2023"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
execute:
  eval: true
  echo: true
  warning: false
---

**EXTRAS**

-   I checked on clean data in 1.3

-   I reflected on the different outliers in 2.1.1

-   I tried different themes on 3.1.1

# 1. Getting Started

## 1.1 Installing and loading required libraries

The code chunk below uses p_load to check if below packages have been installed.

-   **ggrepel**: an R package provides geoms for ggplot2 to repel overlapping text labels.

-   **ggthemes**: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.

-   **hrbrthemes**: an R package provides typography-centric themes and theme components for ggplot2.

-   **patchwork**: an R package for preparing composite figure created using ggplot2.

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse)
```

## 1.2 Importing data

Below code chunk imports a data file called *exam_data*. It consists of year and end examination grades of a cohort of primary 3 students from a local school.

```{r}
exam_data <- read_csv("data/Exam_data.csv", show_col_types = FALSE)
```

## 1.3 Checking data

::: callout-note
## Clean Data?

Only been a week since the last exercise and I am intrigued by what VAA has to offer! After reading a few articles, I found some ways to summarize and check the data. I will try out on this Hands-on Exercise 02's data set and see if we can find anything interesting.
:::

::: panel-tabset
## Summary

```{r}
summary(exam_data)
```

## Structure

```{r}
str(exam_data)
```

## Missing values

```{r}
colSums(is.na(exam_data))
anyNA(exam_data)
```

## Duplicates

```{r}
sum(duplicated(exam_data))
exam_data[duplicated(exam_data), ]
```
:::

::: callout-tip
## Conclusion

We can see that there are no missing values or duplicates. We can conlude data is clean and ready to proceed with plotting!
:::

# 2. Beyond ggplot2 Annotation: ggrepel

The issue with plotting statistical graph is annotation, especially with large number of data points - it would look cluttered!

::: panel-tabset
## The plot

```{r}
#| echo: false
  ggplot(data=exam_data, 
         aes(x= MATHS, 
             y=ENGLISH)) +
    geom_point() +
    geom_smooth(method=lm, 
                size=0.5) +  
    geom_label(aes(label = ID), 
               hjust = .5, 
               vjust = -.5) +
    coord_cartesian(xlim=c(0,100),
                    ylim=c(0,100)) +
    ggtitle("English scores versus Maths scores for Primary 3")

```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## 2.1 Working with ggrepel

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

### 2.1.1 Labelling Outliers

::: callout-note
## Outliers

I was thinking, what if we only wanted to see students who are the outliers? I set the threshold to be \>20 to show students that have scores between English and Maths being more than 20. This can help focus on unusual cases where students excel in one subject, but struggle in the other.
:::

::: panel-tabset
## The plot

```{r}
#| echo: false
exam_data$outlier <- abs(exam_data$ENGLISH - exam_data$MATHS) > 20

ggplot(data = exam_data, 
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = "lm", size = 0.5) +
  geom_text_repel(data = subset(exam_data, outlier == TRUE),
                  aes(label = ID),
                  size = 3) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
exam_data$outlier <- abs(exam_data$ENGLISH - exam_data$MATHS) > 20

ggplot(data = exam_data, 
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = "lm", size = 0.5) +
  geom_text_repel(data = subset(exam_data, outlier == TRUE),
                  aes(label = ID),
                  size = 3) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

::: callout-note
## High-performing vs Underperforming students?

What about students who deviate more than the general average? I applied the simple concept of mean and standard deviation. In below case, if student deviates more than 1 SD from the mean, they are labelled instead.
:::

::: panel-tabset
## The plot

```{r}
#| echo: false
# Calculate score difference for each student
exam_data$score_diff <- exam_data$ENGLISH - exam_data$MATHS

# Compute the mean and standard deviation of the differences
mean_diff <- mean(exam_data$score_diff)
sd_diff <- sd(exam_data$score_diff)

# Flag as outlier if the score difference deviates more than 1 SD from mean
exam_data$outlier <- abs(exam_data$score_diff - mean_diff) > sd_diff

# Plot
ggplot(data = exam_data, 
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = "lm", size = 0.5) +
  geom_text_repel(data = subset(exam_data, outlier == TRUE),
                  aes(label = ID),
                  size = 3) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English vs Maths: Highlighting Students Who Deviate from Class Average")
```

## The code

```{r}
#| eval: false
# Score difference for each student
exam_data$score_diff <- exam_data$ENGLISH - exam_data$MATHS

# Computing the mean and standard deviation of the differences
mean_diff <- mean(exam_data$score_diff)
sd_diff <- sd(exam_data$score_diff)

# Flag as outlier if the score difference deviates more than 1 SD from mean
exam_data$outlier <- abs(exam_data$score_diff - mean_diff) > sd_diff

# Plot
ggplot(data = exam_data, 
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = "lm", size = 0.5) +
  geom_text_repel(data = subset(exam_data, outlier == TRUE),
                  aes(label = ID),
                  size = 3) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English vs Maths: Highlighting Students Who Deviate from Class Average")
```
:::

::: callout-tip
## Findings

It looks like ggrepel is smart enough to know when there's too many overlaps. In above example, it skipped 50 students to keep things readable. Interesting finding! This tells me that it is no different than the initial issue we had when not using ggrepel. I could perhaps increase overlaps by defining `max.overlaps` , but that would still clutter the plot.
:::

# 3. Beyond ggplot2 Themes

ggplot2 comes with eight built-in themes, they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")  
```
:::

## 3.1 Working with ggtheme package

ggthemes provides ‘ggplot2’ themes that replicate the look of plots by Edward Tufte, Stephen Few, Fivethirtyeight, The Economist, ‘Stata’, ‘Excel’, and The Wall Street Journal, among others.

In the example below, The Economist theme is used

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist() 
```
:::

### 3.1.1 Trying other themes

::: callout-note
## Curiosity!

I wondered how other themes look like, so I used the scatter plot from Hands-on Exercise 01 and experimented with the themes!
:::

::: panel-tabset
## Fivethirtyeight

```{r}
ggplot(data = exam_data, 
       aes(x = MATHS, 
           y = ENGLISH, 
           color = GENDER)) +  # Add color aesthetic
  geom_point() +
  theme_fivethirtyeight() +
  ggtitle("English and Maths Scores by Gender")
```

## Solarized

```{r}
ggplot(data = exam_data, 
       aes(x = MATHS, 
           y = ENGLISH, 
           color = GENDER)) +  
  geom_point() +
  theme_solarized() +
  ggtitle("English and Maths scores by Gender")
```

## Few

```{r}
ggplot(data = exam_data, 
       aes(x = MATHS, 
           y = ENGLISH, 
           color = GENDER)) +  
  geom_point() +
  theme_few() +
  ggtitle("English and Maths scores by Gender")
```

## Wall Street Journal

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_wsj() +
  theme(plot.title = element_text(size = 25, face = "bold"))
```
:::

## 3.2 Working with hrbthems package

hrbrthemes package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
:::

The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

::: callout-important
## What can we learn from the code chunk above?

-   `axis_title_size` argument is used to increase the font size of the axis title to 18,

-   `base_size` argument is used to increase the default axis label to 15, and

-   `grid` argument is used to remove the x-axis grid lines.
:::

# 4. Beyond Single Graph

Below is an example of how to create a composite plot by combining multiple graphs. We will first create three statistical graphics by using the code chunk below.

::: panel-tabset
## The plot

```{r}
#| echo: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

print(p1)
```

## The code

```{r}
#| eval: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

print(p1)
```
:::

Next up,

::: panel-tabset
## The plot

```{r}
#| echo: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

print(p2)
```

## The code

```{r}
#| eval: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

print(p2)
```
:::

::: panel-tabset
Lastly, we will draw a scatterplot for English score versus Maths score as shown below

## The plot

```{r}
#| echo: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores vs Maths for Primary 3") 

print(p3)
```

## The code

```{r}
#| eval: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores vs Maths for Primary 3") 

print(p3)
```
:::

## 4.1 Creating Composite Graphics: pathwork methods

Patchwork package has a very simple syntax where we can create layouts super easily. Here’s the general syntax that combines:

-   Two-Column Layout using the Plus Sign +.

-   Parenthesis () to create a subplot group.

-   Two-Row Layout using the Division Sign /

### 4.1.1 Combining two ggplot2 graphs

::: panel-tabset
## The plot

```{r}
#| echo: false
p1 + p2
```

## The code

```{r}
#| eval: false
p1 + p2
```
:::

### 4.1.2 Combining three ggplot2 graphs

Plotting more complex composite using appropriate operations.

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

::: panel-tabset
## The plot

```{r}
#| echo: false
(p1/p2) | p3
```

## The code

```{r}
#| eval: false
(p1/p2) | p3
```
:::

### 4.1.3 Creating figure with insert

Using `inset_element()` of patchwork, we can place one or several plots or graphic elements freely on top or below another plot.

::: panel-tabset
## The plot

```{r}
#| echo: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

## The code

```{r}
#| eval: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```
:::

### 4.1.4 Creating a composite figure by using patchwork and ggtheme

Figure below is created by combining patchwork and theme-economist() of ggthemes package.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 7
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
:::
